---
title: "PARADYS Proteomics analysis"
author: "Fernando Garrido Muñoz"
date: "`r format(Sys.time(), '%a %d, %b, %Y')`"
output: html_document
---

# Load required packages

```{r, warning=FALSE, results='hide', message=FALSE}

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
library(tidyverse)
library(DEqMS)
library(factoextra)
library(FactoMineR)
library(stringr)
library(readxl)
library(writexl)
library(VIM)
library(promor)
library(pcaMethods)
library(ggplot2)
library(gplots)
library(mixOmics)
library(VennDetail)
library(kableExtra)
library(missForest)
library(gtsummary)
library(NormalyzerDE)
library(limma)
library(pROC)


```

# Load and pre-process proteomic data

```{r}

data <- read_excel("NPA_OPT_250424.xlsx")
data <- as.data.frame(data)

```

## Filter rows which contain NAs

```{r}

# Remove duplicate rows based on Protein.IDs:(no duplicated rows)

data <- data[!duplicated(data$Protein.Ids), ]

# Remove rows with NAs protein IDs

data <- data[!is.na(data$Protein.Ids),]

```

## Rework data frame to select our samples with all protein ids (rows):

```{r}

sample_columns = seq(6,length(data),1)
df_protein <- data[, sample_columns]
rownames(df_protein) <- data$Protein.Ids
df_protein[,c(1:131)] <- apply(df_protein[,c(1:131)],2,
                        function(x) as.numeric(as.character(x)))

# We must extract only the samples we need 

# Which is the % of NAs?

sum(is.na(df_protein)/prod(dim(df_protein))) *100

# 40.84 %

```

## Order columns in groups (control vs BPD):

### ALL samples

```{r}

samples <- read_excel("inventario_post_procesamiento.xlsx", sheet = "SamplesConditions")
samples <- as.data.frame(samples)
samples$jensen_condition <- as.factor(samples$jensen_condition) # turn this data to factor
# class(samples$jensen_condition) this is used to check if it has correctly changed to factor
samples$type <- as.factor(samples$type)
# order_samples <- samples$jensen_condition
rownames(samples) <- samples$new_sample

# Add a more specific column with condition + type:

samples <- samples %>% mutate(cond_type=case_when(jensen_condition=="control" & type=="NF" ~ "No Dec",
                                       jensen_condition=="control" & type=="TQ" ~ "TA Control",
                                       jensen_condition=="bpd" & type=="NF" ~ "No Dec",
                                       jensen_condition=="bpd" & type=="TQ" ~ "TA BPD",
                                       jensen_condition=="fallecido" & type=="NF" ~ "Dec",
                                       jensen_condition=="fallecido" & type=="TQ" ~ "TA Dec"),
                              cod_centro=case_when(hospital_abrev=="Cádiz" ~ "01",
                                                   hospital_abrev=="Basurto" ~ "02",
                                                   hospital_abrev=="Vigo" ~ "03",
                                                   hospital_abrev=="Girona" ~ "04",
                                                   hospital_abrev=="Clinic" ~ "05",
                                                   hospital_abrev=="HGM" ~ "06",
                                                   hospital_abrev=="León" ~ "07"))

samples$cond_type <- as.factor(samples$cond_type)
samples <- samples[samples$new_sample %in% colnames(df_protein),]

# Before order, we must keep only the samples we'll need in this experiment:

df_new_prot <- df_protein[,colnames(df_protein) %in% samples$new_sample]
order <- samples$new_samples
df_new_prot <- df_new_prot[,order]

```

## Lets work only with the first protein ID (and generate a file):

```{r}

unique_prot <- df_new_prot
unique_proteins <- gsub(";.*", "", rownames(unique_prot))
data <- data %>%
  mutate(Unique_protein= gsub(";.*", "", rownames(unique_prot)),
         Unique_genes=gsub(";.*", "", Protein.Names))
unique_prot <- unique_prot[!duplicated(unique_proteins), ]
rownames(unique_prot) <- gsub(";.*", "", rownames(unique_prot))
write.table(x = unique_prot,file = "raw_proteins_unique_ord.txt",
            quote = F,row.names = T,
            sep = "\t")

```

## Filter NAs: only filter rows which have more than 30% of NAs

```{r}

noNA_prot <- unique_prot[!rowSums(is.na(unique_prot)),]

# 312 proteins left

unique_prot_ok <- unique_prot[!rowSums(is.na(unique_prot)) 
                                             > ncol(unique_prot)*.3,]

# 1626 proteins left

# Percentage of NAs left:

sum(is.na(unique_prot_ok)/prod(dim(unique_prot_ok))) *100 # 9.85% NAs


```

## Filter our original data for further analysis: 

```{r}

filtered_data <- data %>% filter(data$Protein.Ids %in% rownames(unique_prot_ok))
protein.ids <- filtered_data$Protein.Ids

```

## Impute data using KNN

```{r, eval=FALSE}

prot_knn <- t(unique_prot_ok)
prot_knn <- kNN(prot_knn, k=2, imp_var = F)
prot_knn <- t(prot_knn)
prot_knn <- as.data.frame(prot_knn)
colnames(prot_knn) <- colnames(unique_prot_ok)
write.table(prot_knn, file = "data_knn_all.txt", sep = "\t", 
            col.names = colnames(prot_knn))

```

## log2 transformation:

```{r, results='hide'}

data.knn.all <-read.table(file = "data_knn_all.txt", sep = "\t", header = T,
                      row.names = 1, check.names = F)
# sapply(data.knn, class) Numeric class checker
write.table(x = data.knn.all,file = "Metadata/data_knn_all.tsv",
            quote = F,row.names = F,
            sep = "\t")
protein_log <- log2(data.knn.all)
protein_log_filter <- protein_log %>% filter_all(all_vars(.<=20))

```

# NF samples analysis (with DEP)

## Set limma levels:

```{r}

samples <- samples %>%
  arrange(cond_type) %>%
  mutate(limma_groups=case_when(cond_type=="No Dec" ~ 0,
                                cond_type=="Dec" ~ 1))
```

## Exploratory analysis

### PCA 

#### Required data

```{r}

dim(protein_log_filter)
protein_log_filter <- protein_log_filter[,samples$new_samples]
protein_matrix <- t(protein_log_filter)
dim(protein_matrix)
protein_matrix <- scale(protein_matrix, center = FALSE, scale = TRUE) 
conditions <- samples$cond_type
conditions <- factor(conditions, levels=c("No Dec", "Dec"))
summary(conditions)

```

```{r}

set.seed(222)

pca <- pca(protein_matrix,ncomp = 2, center = TRUE, scale = TRUE,
           max.iter = 100)

plotIndiv(pca, comp = c(1, 2), ind.names = F, 
          group = conditions,  
          legend = TRUE, title = 'PCA comp 1 - 2')

svglite::svglite("FIGURAS/PCA.svg")
plotIndiv(pca, comp = c(1, 2), ind.names = F, 
          group = conditions,  
          legend = TRUE, title = 'PCA comp 1 - 2',
          col.per.group =  c("#009E73","#D55E00"))
dev.off()
```

### PLS-DA

#### Number of components for PLS-DA

```{r}

plsda.perf <- plsda(protein_matrix, conditions, ncomp = 10) 
perf.plsda <- mixOmics::perf(plsda.perf, validation = 'Mfold', folds = 5,
                             progressBar = FALSE, nrepeat = 100) 
svglite::svglite("FIGURAS/NCOMP_PLSDA.svg")
plot(perf.plsda, overlay = 'measure', sd=TRUE)
dev.off()
perf.plsda$choice.ncomp

```

#### Best PLS-DA

```{r}

plsda <- plsda(protein_matrix, conditions, ncomp = 2) # Restrict to 2 components 
plotVar(plsda, comp = c(1,2), cex = 2) 

```

#### Plot PLS-DA

```{r}

# ellipse= FALSE , if you want to turn on the ellipses, just enter ellipse=TRUE

plotIndiv(plsda , comp = c(1,2),
          group = conditions, ind.names = FALSE, 
          ellipse = F, legend = TRUE, title = 'PLSDA comp 1 - 2')

svglite::svglite("FIGURAS/PLSDA.svg")
plotIndiv(plsda , comp = c(1,2),
          group = conditions, ind.names = FALSE, 
          ellipse = F, legend = TRUE, title = 'PLSDA comp 1 - 2',
          col.per.group =  c("#009E73","#D55E00"))
dev.off()

```

### sPLS-DA

```{r}

protein.splsda <- splsda(protein_matrix, conditions, ncomp = 10) 
# set ncomp to 10 for performance assessment later

```

#### Tuning sPLS-DA

```{r}
set.seed(222)
# Tuning
perf.splsda.protein <- perf(protein.splsda, validation = "Mfold", 
                          folds = 5, nrepeat = 100, # Cross Validation
                          progressBar = FALSE, auc = TRUE) 

plot(perf.splsda.protein, col = color.mixo(5:7), sd = TRUE,
     legend.position = "horizontal")
svglite::svglite("FIGURAS/sPLSDA_ncomp.svg")
plot(perf.splsda.protein, col = color.mixo(5:7), sd = TRUE,
     legend.position = "horizontal")
dev.off()


perf.splsda.protein$choice.ncomp # Optimal number of components


```

#### Selecting the number of variables

```{r}
set.seed(222)

list.keepX <- c(1:10,  seq(20, 300, 10))

tune.splsda.protein <- tune.splsda(protein_matrix, conditions, ncomp = 3,
                                 validation = 'Mfold',
                                 folds = 5, nrepeat = 100,
                                 dist = 'mahalanobis.dist', 
                                 measure = "BER", 
                                 test.keepX = list.keepX,
                                 cpus = 4,
                                 progressBar = FALSE) 
plot(tune.splsda.protein, col = color.jet(3)) 
svglite::svglite("FIGURAS/tuned_sPLSDA.svg")
plot(tune.splsda.protein, col = color.jet(3))
dev.off()
tune.splsda.protein$choice.ncomp$ncomp # Optimal number of components
tune.splsda.protein$choice.keepX # Optimal variables for each component
optimal.ncomp <- tune.splsda.protein$choice.ncomp$ncomp
optimal.keepX <- tune.splsda.protein$choice.keepX[1:optimal.ncomp]


```

#### Final sPLS-DA

```{r}
set.seed(222)
final.splsda <- splsda(protein_matrix, conditions, 
                       ncomp = 3,
                       keepX = optimal.keepX) 

```

##### Individual Plots

```{r, warning=FALSE}

plotIndiv(final.splsda, comp = c(1,2), 
          group = conditions, ind.names = FALSE, 
          ellipse = FALSE, legend = TRUE, 
          title = ' (a) sPLS-DA on proteins, comp 1 & 2',
          col.per.group =  c("#009E73","#D55E00"))

background = background.predict(final.splsda, comp.predicted=2, dist = "mahalanobis.dist")

plotIndiv(final.splsda, comp = 1:2,
          group = conditions, ind.names = FALSE, 
          background = background,
          legend = TRUE, title = " (b) sPLSDA with prediction background")

svglite::svglite("FIGURAS/sPLSDA_pred_bg.svg")
plotIndiv(final.splsda, comp = 1:2,
          group = conditions, ind.names = FALSE, 
          background = background,
          legend = TRUE, title = "sPLSDA with prediction background",
          col.per.group =  c("#009E73","#D55E00"))
dev.off()

```

### Prediction and ROC curves

```{r}
set.seed(222)
train <- sample(1:nrow(protein_matrix), 50)
test <- setdiff(1:nrow(protein_matrix), train) 

# Training and test data creation:

X.train <- protein_matrix[train, ]
X.test <- protein_matrix[test,]
Y.train <- conditions[train]
Y.test <- conditions[test]

# Training

train.splsda.protein <- splsda(X.train, Y.train, ncomp = 3, keepX = optimal.keepX) 

# Test the model

predict.splsda.protein <- predict(train.splsda.protein, X.test, 
                                dist = "mahalanobis.dist")

# Predictive Accuracy Comp 1:

predict.comp1 <- predict.splsda.protein$class$mahalanobis.dist[,1]
table(factor(predict.comp1, levels = levels(conditions)), Y.test)

# Predictive Accuracy Comps 1 & 2:

predict.comp2 <- predict.splsda.protein$class$mahalanobis.dist[,2]
table(factor(predict.comp2, levels = levels(conditions)), Y.test)

```

#### Plot ROC curve

```{r}

svglite::svglite("FIGURAS/ROC_1_comp.svg")
auc.splsda = auroc(final.splsda, roc.comp = 1, print = TRUE) 
dev.off()
svglite::svglite("FIGURAS/ROC_2_comp.svg")
auc.splsda2 = auroc(final.splsda, roc.comp = 2, print = TRUE)
dev.off()

```

#### CI ROC curve

```{r}

pred.final.splsda <- predict(final.splsda, protein_matrix, 
                                dist = "mahalanobis.dist")
DiscriminantScoresTable.dim1 <- cbind(as.data.frame(conditions) , as.data.frame(pred.final.splsda$predict[,,"dim1"])) #the number in dimensions depends on the number of components included in your model 
DiscriminantScoresTable.dim2 <- cbind(as.data.frame(conditions) , as.data.frame(pred.final.splsda$predict[,,"dim2"])) #the number in dimensions depends on the number of components included in your model 
DiscriminantScoresTable.dim3 <- cbind(as.data.frame(conditions) , as.data.frame(pred.final.splsda$predict[,,"dim3"]))


```

```{r}
library(pROC)

ROC.dim1 <- roc(DiscriminantScoresTable.dim1$conditions, 
                DiscriminantScoresTable.dim1$Dec) 

ROC.dim1$auc # AUC

ci.auc(ROC.dim1$auc) # 95% CI

ROC.dim2 <- roc(DiscriminantScoresTable.dim2$conditions, 
                DiscriminantScoresTable.dim2$Dec) 

ROC.dim2$auc # AUC

ci.auc(ROC.dim2$auc) # 95% CI

ROC.dim3 <- roc(DiscriminantScoresTable.dim3$conditions, 
                DiscriminantScoresTable.dim3$Dec) 

ROC.dim3$auc # AUC

ci.auc(ROC.dim3$auc) # 95% CI


```


## Protein Differential Expression Analysis

Using limma

Contrasts to perform: 

+ No Dec vs Dec

### Prepare our data

```{r}

experimental.design <- model.matrix(~ -1+factor(samples$limma_groups)) 
colnames(experimental.design) <- c("No_Dec","Dec")
linear.fit <- lmFit(protein_log_filter, experimental.design)
contrast.matrix <- makeContrasts(Dec-No_Dec,
                                 levels=c("No_Dec","Dec"))

```

### P-values and Fold-Change estimation using *contrast.fit* and *eBayes*

```{r}

contrast.linear.fit <- contrasts.fit(linear.fit, contrast.matrix)
contrast.results <- eBayes(contrast.linear.fit)

```

### Contrasts

#### No Dec vs Dec

```{r}

control.bpd <- topTable(contrast.results, number=nrow(protein_log_filter),
                           coef=1,sort.by="logFC")
head(control.bpd)

log.fold.change <- control.bpd$logFC
p.value <- control.bpd$P.Value
q.value <- control.bpd$adj.P.Val
protein.ids <- rownames(control.bpd)
names(log.fold.change) <- protein.ids
names(p.value) <- protein.ids

dep <- protein.ids[abs(log.fold.change) > 0.58 & p.value < 0.05]
up.proteins <- protein.ids[log.fold.change > 0.58 & p.value < 0.05]
down.proteins <- protein.ids[log.fold.change < -0.58 & p.value < 0.05]

length(up.proteins)
length(down.proteins)

```
##### Save results

```{r, eval=FALSE}

write.table(dep, file= "NPA_filter/dep_nf.txt", sep = "\t", row.names = F, col.names = F)
write.table(up.proteins, file= "NPA_filter/up_dep_nf.txt", sep = "\t",row.names = F, col.names = F)
write.table(down.proteins, file= "NPA_filter/down_dep_nf.txt", sep = "\t",row.names = F, col.names = F)

```

## Analysis of selected proteins after STRING Functional Enrichment Analysis

### Obtention of protein names and additional DEP results table

```{r}

protein.names <- filtered_data$First.Protein.Description[filtered_data$Unique_protein %in%
                                                           rownames(protein_log_filter[dep,])]
protein.names <- c(protein.names, "Unknown_1", "Unknown_2","Unknown_3",
                   "Unknown_4","Unknown_5","Unknown_6")

id.names <- read_excel("idmapping_2024_07_29.xlsx", col_names = TRUE)
id.names <- id.names %>% mutate(`protein name` = gsub("\\s.*", "", id.names$`Gene Names`))

clusters.str <- read_excel("./STRING/string_kmeans_clusters.xlsx")
clusters.str <- as.data.frame(clusters.str)
clusters.str$`protein name`[11] <- "DEFA1"
clusters.str$`protein name`[13] <- "GBA1"
clusters.str$`protein name`[15] <- "H3C1"
clusters.table <- merge(id.names, clusters.str, by = "protein name")
clusters.table <- clusters.table[,-c(3:4, 6:10, 14)]
control.bpd.names <- control.bpd[dep,]
rownames(control.bpd.names) <- id.names$`Entry Name`

# Table with relevant information of DEP: 

control.bpd.names.info <- control.bpd.names
control.bpd.names.info$`Entry Name` <- rownames(control.bpd.names.info)
dep.info.table <- merge(control.bpd.names.info, id.names, by="Entry Name")
dep.info.table <- dep.info.table[,c(7,1,14,9:10,2:6)]
colnames(dep.info.table)[1:3] <- c("Protein ID", "Uniprot Name", "Gene Name")
write_xlsx(dep.info.table, "DEP_results.xlsx")

# Selected proteins table

prot.names.table <- control.bpd.names[clusters.table$`Entry Name`,]
rownames(prot.names.table) <- clusters.table$`protein name`
prot.names.table <- prot.names.table %>%
  arrange(desc(abs(logFC)))
prot.names.table$`protein name` <- rownames(prot.names.table)
final.cluster.table <- merge(clusters.table, prot.names.table, by = "protein name")
final.cluster.table <- final.cluster.table[,-c(5:6,9)]
final.cluster.table <- final.cluster.table %>%
  arrange(desc(logFC))
cluster.1 <- final.cluster.table[final.cluster.table$`cluster number`==1,]
cluster.2 <- final.cluster.table[final.cluster.table$`cluster number`==2,]
cluster.3 <- final.cluster.table[final.cluster.table$`cluster number`==3,]
cluster.4 <- final.cluster.table[final.cluster.table$`cluster number`==4,]

```

### Table design for box and violin plotting, and further analyses

```{r}

selected.final.proteins <- c(final.cluster.table$`Entry Name`)
conditions <- samples$cond_type
conditions <- factor(conditions, levels = c("Dec", "No Dec"))

```

### Calculate AUC for all selected proteins

```{r}

roc.data.sel <- protein_log_filter[dep,]
roc.data.sel <- t(roc.data.sel)
roc.data.sel <- as.data.frame(roc.data.sel)
colnames(roc.data.sel) <- id.names$`Entry Name`
roc.data.sel <- roc.data.sel[,selected.final.proteins]
samples.roc <- samples
samples.roc$class <- samples.roc$cond_type
samples.roc$class <- factor(samples.roc$class, levels=c("No Dec", "Dec"), 
                         labels=c("0", "1"))
roc.data.sel <- roc.data.sel[samples.roc$new_samples,]

auc.df <- data.frame(proteins= colnames(roc.data.sel), AUC=NA)
for (n in 1:nrow(auc.df)){
  name <- auc.df$proteins[n]
  h <- roc(samples.roc$class, roc.data.sel[,name], plot=FALSE, legacy.axes=TRUE, print.auc=TRUE,print.auc.x=.3,print.auc.y=0.53, grid=TRUE)
  h1 <- h[["auc"]][1]
  auc.df$AUC[n] <- h1
}

auc.df <- auc.df %>%
  arrange(desc(AUC))

glm.fit.panel.sel.5 <- glm(samples.roc$class ~ roc.data.sel[,auc.df$proteins[1]] +
                           roc.data.sel[,auc.df$proteins[2]] +
                           roc.data.sel[,auc.df$proteins[3]] +
                           roc.data.sel[,auc.df$proteins[4]] + 
                           roc.data.sel[,auc.df$proteins[5]],
                         family=binomial)
glm.fit.panel.sel.10 <- glm(samples.roc$class ~ roc.data.sel[,auc.df$proteins[1]] +
                           roc.data.sel[,auc.df$proteins[2]] +
                           roc.data.sel[,auc.df$proteins[3]] +
                           roc.data.sel[,auc.df$proteins[4]] + 
                           roc.data.sel[,auc.df$proteins[5]] +
                           roc.data.sel[,auc.df$proteins[6]] +
                           roc.data.sel[,auc.df$proteins[7]] +
                           roc.data.sel[,auc.df$proteins[8]] +
                           roc.data.sel[,auc.df$proteins[9]] +
                           roc.data.sel[,auc.df$proteins[10]],
                         family=binomial)
glm.fit.panel.sel.15 <- glm(samples.roc$class ~ roc.data.sel[,auc.df$proteins[1]] +
                           roc.data.sel[,auc.df$proteins[2]] +
                           roc.data.sel[,auc.df$proteins[3]] +
                           roc.data.sel[,auc.df$proteins[4]] + 
                           roc.data.sel[,auc.df$proteins[5]] +
                           roc.data.sel[,auc.df$proteins[6]] +
                           roc.data.sel[,auc.df$proteins[7]] +
                           roc.data.sel[,auc.df$proteins[8]] +
                           roc.data.sel[,auc.df$proteins[9]] +
                           roc.data.sel[,auc.df$proteins[10]] +
                           roc.data.sel[,auc.df$proteins[11]] +
                           roc.data.sel[,auc.df$proteins[12]] +
                           roc.data.sel[,auc.df$proteins[13]] +
                           roc.data.sel[,auc.df$proteins[14]] +
                           roc.data.sel[,auc.df$proteins[15]],
                         family=binomial)
glm.fit.panel.sel.20 <- glm(samples.roc$class ~ roc.data.sel[,auc.df$proteins[1]] +
                           roc.data.sel[,auc.df$proteins[2]] +
                           roc.data.sel[,auc.df$proteins[3]] +
                           roc.data.sel[,auc.df$proteins[4]] + 
                           roc.data.sel[,auc.df$proteins[5]] +
                           roc.data.sel[,auc.df$proteins[6]] +
                           roc.data.sel[,auc.df$proteins[7]] +
                           roc.data.sel[,auc.df$proteins[8]] +
                           roc.data.sel[,auc.df$proteins[9]] +
                           roc.data.sel[,auc.df$proteins[10]] +
                           roc.data.sel[,auc.df$proteins[11]] +
                           roc.data.sel[,auc.df$proteins[12]] +
                           roc.data.sel[,auc.df$proteins[13]] +
                           roc.data.sel[,auc.df$proteins[14]] +
                           roc.data.sel[,auc.df$proteins[15]] +
                           roc.data.sel[,auc.df$proteins[16]] +
                           roc.data.sel[,auc.df$proteins[17]] +
                           roc.data.sel[,auc.df$proteins[18]] +
                           roc.data.sel[,auc.df$proteins[19]] +
                           roc.data.sel[,auc.df$proteins[20]],
                         family=binomial)

glm.fit.panel.sel.sig <- glm(samples.roc$class ~
                           roc.data.sel[,auc.df$proteins[1]] +
                           roc.data.sel[,auc.df$proteins[2]] + 
                           roc.data.sel[,auc.df$proteins[3]] +
                           roc.data.sel[,auc.df$proteins[4]] +
                           roc.data.sel[,auc.df$proteins[5]] +
                           roc.data.sel[,auc.df$proteins[11]] +
                           roc.data.sel[,auc.df$proteins[14]],
                         family=binomial)

glm.fit.panel.sel.sig.top <- glm(samples.roc$class ~
                           roc.data.sel[,auc.df$proteins[2]] +   
                           roc.data.sel[,auc.df$proteins[3]] +
                           roc.data.sel[,auc.df$proteins[4]] +
                           roc.data.sel[,auc.df$proteins[5]],
                         family=binomial)

glm.fit.panel.sel.sig.top3<- glm(samples.roc$class ~
                           roc.data.sel[,auc.df$proteins[2]] +   
                           roc.data.sel[,auc.df$proteins[4]] +
                           roc.data.sel[,auc.df$proteins[5]],
                         family=binomial)

par(pty="s")
roc(samples.roc$class, glm.fit.panel.sel.5$fitted.values, plot=TRUE,
    legacy.axes=TRUE,print.auc=TRUE,print.auc.x=.65,print.auc.y=0.53, grid=TRUE,
    col="#D81B60", lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.10$fitted.values, col="#1E88E5", add=TRUE,
         print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.43, lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.15$fitted.values, col="#FFC107", add=TRUE,
         print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.33, lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.20$fitted.values, col="#004D40", add=TRUE,
         print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.23, lwd=4, ci=TRUE)
par(pty="m")

par(pty="s")
roc(samples.roc$class, glm.fit.panel.sel.sig$fitted.values, plot=TRUE,
    legacy.axes=TRUE,print.auc=TRUE,print.auc.x=.65,print.auc.y=0.53, grid=TRUE,
    col="#A727BD", lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.sig.top$fitted.values, col="#ADE718",
         add=TRUE, print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.43, lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.sig.top3$fitted.values, col="#5BD5FA",
         add=TRUE, print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.33, lwd=4, ci=TRUE)
par(pty="m")

summary(glm.fit.panel.sel.5)
summary(glm.fit.panel.sel.10)
summary(glm.fit.panel.sel.15)
summary(glm.fit.panel.sel.20)
summary(glm.fit.panel.sel.sig)
summary(glm.fit.panel.sel.sig.top)
summary(glm.fit.panel.sel.sig.top3)

# Get CI values of each selected biomarker, then for all panels

CI.df <- data.frame(proteins= colnames(roc.data.sel[,auc.df$proteins[c(1:5,11,
                                                                       14)]]),
                    CI=NA)
for (n in 1:nrow(CI.df)){
  name <- auc.df$proteins[n]
  h <- roc(samples.roc$class, roc.data.sel[,name], plot=FALSE, legacy.axes=TRUE, print.auc=TRUE,print.auc.x=.3,print.auc.y=0.53, grid=TRUE,ci=TRUE)
  h1 <- paste0(h[["ci"]][1]," - ",h[["ci"]][3])
  CI.df$CI[n] <- h1
}

CI.df

write_xlsx(CI.df, "CI_biomarkers.xlsx")

CI.bm.panels <- data.frame(Biomarker.panels= c("Top 7","Top 4", "Top 3"),
                    CI=NA)
top7bm <- roc(samples.roc$class, glm.fit.panel.sel.sig$fitted.values,  
    print.auc=TRUE,legacy.axes=TRUE, grid=TRUE, ci=TRUE)
top4bm <- roc(samples.roc$class, glm.fit.panel.sel.sig.top$fitted.values, 
    print.auc=TRUE, ci=TRUE)
top3bm <- roc(samples.roc$class, glm.fit.panel.sel.sig.top3$fitted.values,
    print.auc=TRUE, ci=TRUE)

CI.bm.panels$CI[1] <- paste0(top7bm[["ci"]][1]," - ",top7bm[["ci"]][3])
CI.bm.panels$CI[2] <- paste0(top4bm[["ci"]][1]," - ",top4bm[["ci"]][3])
CI.bm.panels$CI[3] <- paste0(top3bm[["ci"]][1]," - ",top3bm[["ci"]][3])

write_xlsx(CI.bm.panels, "CI_biomarker_panels.xlsx")

CI.panels <- data.frame(Panels.AUC= c("Top 5","Top 10", "Top 15", "Top 20"),
                    CI=NA)

top5 <- roc(samples.roc$class, glm.fit.panel.sel.5$fitted.values,
    print.auc=TRUE, ci=TRUE)
top10 <- roc(samples.roc$class, glm.fit.panel.sel.10$fitted.values,
    print.auc=TRUE, ci=TRUE)
top15 <- roc(samples.roc$class, glm.fit.panel.sel.15$fitted.values,
    print.auc=TRUE, ci=TRUE)
top20 <- roc(samples.roc$class, glm.fit.panel.sel.20$fitted.values,
    print.auc=TRUE, ci=TRUE)

CI.panels$CI[1] <- paste0(top5[["ci"]][1]," - ",top5[["ci"]][3])
CI.panels$CI[2] <- paste0(top10[["ci"]][1]," - ",top10[["ci"]][3])
CI.panels$CI[3] <- paste0(top15[["ci"]][1]," - ",top15[["ci"]][3])
CI.panels$CI[4] <- paste0(top20[["ci"]][1]," - ",top20[["ci"]][3])

write_xlsx(CI.panels, "CI_AUC_panels.xlsx")

```

#### Roc TOP comparisons (tests)

```{r}

top5.roc <- roc(samples.roc$class, glm.fit.panel.sel.5$fitted.values, plot=F,
    legacy.axes=TRUE,print.auc=TRUE,print.auc.x=.4,print.auc.y=0.53, grid=TRUE,
    col="red", lwd=4)
top10.roc <- roc(samples.roc$class, glm.fit.panel.sel.10$fitted.values, plot=F,
    legacy.axes=TRUE,print.auc=TRUE,print.auc.x=.4,print.auc.y=0.53, grid=TRUE,
    col="red", lwd=4)
top15.roc <- roc(samples.roc$class, glm.fit.panel.sel.15$fitted.values, plot=F,
    legacy.axes=TRUE,print.auc=TRUE,print.auc.x=.4,print.auc.y=0.53, grid=TRUE,
    col="red", lwd=4)
top20.roc <- roc(samples.roc$class, glm.fit.panel.sel.20$fitted.values, plot=F,
    legacy.axes=TRUE,print.auc=TRUE,print.auc.x=.4,print.auc.y=0.53, grid=TRUE,
    col="red", lwd=4)

roc.test(top5.roc, top10.roc, method="delong")
roc.test(top5.roc, top15.roc, method="delong")
roc.test(top5.roc, top20.roc, method="delong")
roc.test(top10.roc, top15.roc, method="delong")
roc.test(top10.roc, top20.roc, method="delong")
roc.test(top15.roc, top20.roc, method="delong")

```

#### Save figures

```{r, eval=FALSE}

svglite::svglite("FIGURAS/ROC_top_5_20.svg")
par(pty="s")
roc(samples.roc$class, glm.fit.panel.sel.5$fitted.values, plot=TRUE,
    legacy.axes=TRUE,print.auc=TRUE,print.auc.x=.65,print.auc.y=0.53, grid=TRUE,
    col="#D81B60", lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.10$fitted.values, col="#1E88E5", add=TRUE,
         print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.43, lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.15$fitted.values, col="#FFC107", add=TRUE,
         print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.33, lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.20$fitted.values, col="#004D40", add=TRUE,
         print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.23, lwd=4, ci=TRUE)
par(pty="m")
dev.off()

svglite::svglite("FIGURAS/ROC_top_3_7.svg")
par(pty="s")
roc(samples.roc$class, glm.fit.panel.sel.sig$fitted.values, plot=TRUE,
    legacy.axes=TRUE,print.auc=TRUE,print.auc.x=.65,print.auc.y=0.53, grid=TRUE,
    col="#A727BD", lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.sig.top$fitted.values, col="#ADE718",
         add=TRUE, print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.43, lwd=4, ci=TRUE)
plot.roc(samples.roc$class, glm.fit.panel.sel.sig.top3$fitted.values, col="#5BD5FA",
         add=TRUE, print.auc=TRUE,
         print.auc.x=.65,print.auc.y=0.33, lwd=4, ci=TRUE)
par(pty="m")
dev.off()

```


### Table with merged information of AUC, FC, p-values and B values:

```{r}

# Decision table building

merged.dec.table <- control.bpd.names[selected.final.proteins,]
merged.dec.table$proteins <- rownames(merged.dec.table)
rownames(merged.dec.table) <- merged.dec.table$proteins
selected.final.genes <- c(final.cluster.table$`protein name`)
merged.dec.table$genes <- selected.final.genes
merged.dec.table <- merge(merged.dec.table, auc.df, by="proteins")
rownames(merged.dec.table) <- merged.dec.table$proteins 
merged.dec.table <- merged.dec.table[selected.final.proteins,]
merged.dec.table$genes <- selected.final.genes
id.names$proteins <- id.names$`Entry Name`
merged.dec.table <- merge(merged.dec.table, id.names, by="proteins")
merged.dec.table <- merged.dec.table[,-c(9,11:16)]
colnames(merged.dec.table)[9] <- "ID"
merged.dec.table <- merged.dec.table %>%
  arrange(desc(AUC))
order <- c(1,7,9,10, 2:6, 8)
merged.dec.table <- merged.dec.table[,order]
write_xlsx(merged.dec.table, "selected_final_proteins.xlsx")

```

### Table with annotations and clusters included of each protein

```{r}

annotations <- read.xlsx("STRING/string_protein_annotations.xlsx")
annotations <- annotations[,c(1,3)]
colnames(annotations) <- c("genes","description")
annotations$genes[38] <- "DEFA1"
annotations$genes[47] <- "GBA1"
annotations$genes[51] <- "H3C1"
annotation.merged.table <- merge(merged.dec.table, annotations, by="genes")
annotation.merged.table <- annotation.merged.table[,c(3,2,1,5,7:8,10:11)]
annotation.merged.table <- annotation.merged.table %>%
  mutate(protein.description= gsub(";.*", "", description),
         ext.description= gsub("^.*?;", "", description))
annotation.merged.table <- annotation.merged.table[,c(1:3,10,9,4:7)]
clusters.string <- final.cluster.table[,c(1,4)]
colnames(clusters.string) <- c("genes", "cluster.number")
cluster.info <- rbind(clusters.string)
annotation.merged.table <- merge(annotation.merged.table, cluster.info, by="genes")
annotation.merged.table <- annotation.merged.table[,c(1:3,5,10,4,6:9)]
annotation.merged.table <- annotation.merged.table[!duplicated(annotation.merged.table$genes),]
annotation.merged.table <- annotation.merged.table %>%
  mutate(cluster.summary=case_when(cluster.number=="1" ~ "Regulation of molecular functions",
                                   cluster.number=="2" ~ "Proteasome; Acetylation",
                                   cluster.number=="3" ~ "Immune system",
                                   cluster.number=="4" ~ "Vesicle proteins")) %>%
  arrange(desc(AUC))
annotation.merged.table <- annotation.merged.table[,c(1:5,11,6,7:10)]

write_xlsx(annotation.merged.table, "final_proteins_annotations.xlsx")

```

#### With boxplots

```{r}

key.prots <- read_xlsx("DEP_results.xlsx")
key.prots <- arrange(key.prots, desc(abs(key.prots$logFC)))
boxes.prot <- protein_log_filter[key.prots$`Protein ID`[1:10],]
rownames(boxes.prot) <- key.prots$`Gene Name`[1:10]
boxes.prot <- as.data.frame(t(boxes.prot))
boxes.prot$condition <- conditions
library(reshape2)
library(ggsignif)
library(ggpubr)
boxes.prot$condition <- factor(boxes.prot$condition, levels = c("No Dec", "Dec"))
df.m <- melt(boxes.prot, id.var = "condition")
boxplots <- ggplot(data = df.m, aes(x=variable, y=value)) +
  ylim(7.5,22) +
  ylab("log2(Raw Intensities)") +
  xlab("Biomarkers") +
  theme_bw()+
  geom_boxplot(aes(fill=condition)) +
  scale_fill_manual(values = c("#009E73","#D55E00"), labels=c("No Dec", "Dec"),
                    name="Legend")
boxplots
ggsave(plot = boxplots, "FIGURAS/Boxplots.svg", height = 7, width = 9)
  
```

### Volcano with gene names representation

```{r}
data.genes <- data
data.genes$Genes <- gsub(";.*", "", data.genes$Genes)
data.genes$Protein.Ids <- gsub(";.*", "", data.genes$Protein.Ids)
volcano <- control.bpd
volcano <- volcano[,c(1,3)]
volcano$Protein.Ids <- rownames(control.bpd)
volcano <- merge(volcano, data.genes, by= "Protein.Ids")
volcano <- volcano[!duplicated(volcano$Protein.Ids),]
volcano <- volcano[,c(6,2,3)]
volcano$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP" 
volcano$diffexpressed[volcano$logFC > 0.58 & volcano$P.Value < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
volcano$diffexpressed[volcano$logFC < -0.58 & volcano$P.Value < 0.05] <- "DOWN"
volcano$delabel <- NA
volcano$delabel[volcano$diffexpressed != "NO"] <- volcano$Genes[volcano$diffexpressed != "NO"]
library(ggrepel)
volcano.plot <- ggplot(data=volcano, aes(x=logFC, y=-log10(P.Value), 
                    col=diffexpressed, label=delabel)) +
  ylab("-log10(P-value)") +
  xlab("log2(Fold-Change)")+
  xlim(-2.5, 2.5) +
  geom_point() + 
  theme_bw() +
  geom_text_repel(size = 2, nudge_x = 0.05, 
                  nudge_y = 0.05, check_overlap = T, 
                  show.legend=FALSE, col="black") +
  scale_color_manual(values=c("#1A85FF","gray54","#D41159"),
                    name=NULL, labels=c("Downregulated",
                                        "Non-differential",
                                        "Upregulated")) +
  geom_vline(xintercept=c(-0.6, 0.6), col="black", lty=2) +
  geom_hline(yintercept=-log10(0.05), col="black", lty=2) +
  annotate(geom = "text", label=paste0("Downregulated: ", length(down.proteins)),
           col= "#1A85FF", x=-1.8, y=14, fontface= "bold") +
  annotate(geom = "text", label=paste0("Upregulated: ", length(up.proteins)),
           col= "#D41159", x=1.9, y=14, fontface= "bold")

volcano.plot
ggsave(plot= volcano.plot, "Figuras/Volcano_plot.svg",height = 6,width = 8)
```


##### Heatmap

```{r}

heat.prots <- protein_log_filter[key.prots$`Protein ID`,]
rownames(heat.prots) <- key.prots$`Gene Name`

# Heatmap 

colfunc <- colorRampPalette(colors = c("royalblue4","ivory", "red4"))

heatmap.2(as.matrix(heat.prots[1:50,]), 
          col = colfunc(256),
          scale = "row",
          trace = "none",
          dendrogram="both",
          Colv = TRUE,
          key = TRUE,
          keysize = 1.5, 
          key.title = NA,
          ColSideColors = c(rep("#D55E00",14),rep("#009E73",117)),
          labRow = rownames(heat.prots)[1:50],
          cexRow = 0.4,
          srtRow = 0.5,
          key.xlab = "Z-score",
          key.ylab = "",
          density.info = "none",
          cex.main = 1,
          offsetRow = -0.5
          )
par(xpd=TRUE)
legend("topright", legend = c("No Dec", "Dec"), fill = c("#009E73","#D55E00"))
```

##### Save:

```{r}

svglite::svglite("FIGURAS/heatmap_top.svg")
heatmap.2(as.matrix(heat.prots[1:50,]), 
          col = colfunc(256),
          scale = "row",
          trace = "none",
          dendrogram="both",
          Colv = TRUE,
          key = TRUE,
          keysize = 1.5, 
          key.title = NA,
          ColSideColors = c(rep("#D55E00",14),rep("#009E73",117)),
          labRow = rownames(heat.prots)[1:50],
          cexRow = 0.4,
          srtRow = 0.5,
          key.xlab = "Z-score",
          key.ylab = "",
          density.info = "none",
          cex.main = 1,
          offsetRow = -0.5
          )
par(xpd=TRUE)
legend("topright", legend = c("No Dec", "Dec"), fill = c("#009E73","#D55E00"))
dev.off()

```


# Study Population variable table

```{r}

variables <- read_xlsx("./pacientes.xlsx")
variables <- variables[,c(3,5:6,8:9,11,54,70)]
variables <- merge(samples, variables, by=c("patient", "cod_centro"))
variables$`Fecha de Nacimiento` <- as.Date(variables$`Fecha de Nacimiento`, "%d/%m/%Y")
variables$`Fecha de alta` <- as.Date(variables$`Fecha de alta`, "%d/%m/%Y")
variables <- variables %>%
  mutate(Condition=case_when(jensen_condition=="bpd" ~ "No Deceased",
                             jensen_condition=="control" ~ "No Deceased",
                             jensen_condition=="fallecido" ~ "Deceased"),
         BPD.Grade=case_when(jensen=="grado 1" ~ "Grade 1",
                             jensen=="grado 2" ~ "Grade 2",
                             jensen=="grado 3" ~ "Grade 3",
                             jensen=="control" ~ "Control",
                             jensen=="fallecido" ~ "Deceased"),
         Days.hospitalization=difftime(`Fecha de alta`,`Fecha de Nacimiento`, 
                                       units = "days"),
         New.Sex=case_when(Sexo=="Mujer" ~ "Yes",
                           Sexo=="Varón" ~ "No"))  %>%
  mutate_all(function(x) gsub("Varón","Male",x)) %>%
  mutate_all(function(x) gsub("Mujer","Female",x))

variables <- variables %>%
  mutate_at(c("EG al nacimiento (semanas)", "Peso al nacimiento", 
              "Days.hospitalization", "Días 1 sem de VM"), as.numeric)
variables$Condition <- factor(variables$Condition, levels=c("No Deceased", 
                                                            "Deceased"))
variables$BPD.Grade <- factor(variables$BPD.Grade, levels=c("Control", "Grade 1",
                                                            "Grade 2", "Grade 3",
                                                            "Deceased"))
variables$New.Sex <- factor(variables$New.Sex, levels = c("Yes", "No"))
library(openxlsx)
wide.modality <- read.xlsx("wide_060324_mod.xlsx")
wide.modality <- wide.modality %>%
  mutate(Modality=case_when(Modalidad=="AC/VG" | Modalidad=="PC" | 
                              Modalidad=="PSV" | Modalidad=="PSV+VG" |
                              Modalidad=="SIMV+VG" | Modalidad=="VAFO" |
                              Modalidad=="VCRP" | Modalidad=="VM A/C" |
                              Modalidad=="VM A/C CON VG" | 
                              Modalidad=="VMC" | Modalidad=="VMC A/C" ~ "IMV",
                            Modalidad=="VNI" | Modalidad=="DUOPAP" |
                              Modalidad=="IPPVn" ~ "NIV",
                            Modalidad=="CPAP" | Modalidad=="CPAPn" ~ "nCPAP",
                            Modalidad=="Alto flujo" | Modalidad=="GNAF" |
                              Modalidad=="OAF" | Modalidad=="HFNC" ~ "HFNC",
                            Modalidad=="GNBF" ~ "LFNC",
                            Modalidad=="No O2" | 
                              Modalidad=="Sin soporte" ~ "No respiratory support"
                            ))
wide.modality$Modality <- factor(wide.modality$Modality, levels = c("IMV", "NIV",
                                                                    "nCPAP",
                                                                    "HFNC","LFNC",
                                                                    "No respiratory support"))
new.variables <- merge(variables, wide.modality, by=c("patient","cod_centro"))
hospitals <- new.variables[,c(18,4)]

new.variables <- new.variables[,c(16,19,12:14,21,17:18)]

# Hospitals

hospitals %>%
  tbl_summary(
    by= Condition,
    statistic = list(hospital_abrev ~ "{n} ({p}%)"),   
    type   = list(hospital_abrev ~ "categorical"),
    label = list(hospital_abrev ~ "Hospital"),
    sort = list(everything() ~ "frequency")) %>%
  modify_header(label ~ "**Condition**") %>%
  modify_caption("**Table. Patient Hospital of Origin**") %>%
  add_overall() %>%
  italicize_levels()

# Hospital save:

hospitals %>%
  tbl_summary(
    by= Condition,
    statistic = list(hospital_abrev ~ "{n} ({p}%)"),   
    type   = list(hospital_abrev ~ "categorical"),
    label = list(hospital_abrev ~ "Hospital"),
    sort = list(everything() ~ "frequency")) %>%
  modify_header(label ~ "**Condition**") %>%
  modify_caption("**Table. Patient Hospital of Origin**") %>%
  add_overall() %>%
  italicize_levels() %>%
  as_gt() %>%
  gt::gtsave(filename = "pacientes_hospitales.docx")

# View table:

new.variables %>%
  tbl_summary(
    by= Condition,
    statistic = list(all_continuous() ~ "{mean} ({sd})",        
                     BPD.Grade ~ "{n} ({p}%)",
                     New.Sex ~ "{n} ({p}%)"),   
    digits = all_continuous() ~ 1,                            
    type   = list(all_categorical() ~ "categorical",
                  `EG al nacimiento (semanas)` ~ "continuous",
                  `Días 1 sem de VM` ~ "continuous",
                  Modality ~ "categorical"),
    label = list(New.Sex ~ "Sex, Female",
                 `EG al nacimiento (semanas)` ~ "GA (weeks)",
                 `Peso al nacimiento` ~ "Weight",
                 `Días 1 sem de VM` ~ "Days of VM at 1 week",
                 BPD.Grade ~ "BPD grade",
                 Days.hospitalization ~ "Days of hospitalization",
                 Modality ~ "Respiratory modality al 1 week")) %>%
  modify_header(label ~ "**Condition**") %>%
  modify_caption("**Table 1. Patient Clinical Characteristics**") %>%
  add_p() %>%
  bold_p() %>%
  add_overall() %>%
  italicize_levels() %>%
  as_gt() %>%
  gt::tab_source_note(gt::md("*IVM: Invasive Mechanical Ventilation; NIV: Non-Invasive Ventilation; nCPAP: nasal CPAP; HFNC: High Flow Nasal Cannula ; LFNG: Low Flow Nasal Cannula*"))

# Save table as .docx

new.variables %>%
  tbl_summary(
    by= Condition,
    statistic = list(all_continuous() ~ "{mean} ({sd})",        
                     BPD.Grade ~ "{n} ({p}%)",
                     New.Sex ~ "{n} ({p}%)"),   
    digits = all_continuous() ~ 1,                            
    type   = list(all_categorical() ~ "categorical",
                  `EG al nacimiento (semanas)` ~ "continuous",
                  `Días 1 sem de VM` ~ "continuous",
                  Modality ~ "categorical"),
    label = list(New.Sex ~ "Sex, Female",
                 `EG al nacimiento (semanas)` ~ "GA (weeks)",
                 `Peso al nacimiento` ~ "Weight",
                 `Días 1 sem de VM` ~ "Days of VM at 1 week",
                 BPD.Grade ~ "BPD grade",
                 Days.hospitalization ~ "Days of hospitalization",
                 Modality ~ "Respiratory modality al 1 week")) %>%
  modify_header(label ~ "**Condition**") %>%
  modify_caption("**Table 1. Patient Clinical Characteristics**") %>%
  add_p() %>%
  bold_p() %>%
  add_overall() %>%
  italicize_levels() %>%
  as_gt() %>%
  gt::tab_source_note(gt::md("*IVM: Invasive Mechanical Ventilation; NIV: Non-Invasive Ventilation; nCPAP: nasal CPAP; HFNC: High Flow Nasal Cannula ; LFNG: Low Flow Nasal Cannula*")) %>%
  gt::gtsave(filename = "Table_1_def_nodec_dec.docx")



```

# Overall table including deceased

```{r}

variables <- read_xlsx("./pacientes.xlsx")
variables <- variables[,c(3,5:6,8:9,11,54,70,73)]
variables <- merge(samples, variables, by=c("patient", "cod_centro"))
variables$`Fecha de Nacimiento` <- as.Date(variables$`Fecha de Nacimiento`, "%d/%m/%Y")
variables$`Fecha de alta` <- as.Date(variables$`Fecha de alta`, "%d/%m/%Y")
variables <- variables %>%
  mutate(Condition=case_when(jensen_condition=="bpd" ~ "BPD",
                             jensen_condition=="control" ~ "Control",
                             jensen_condition=="fallecido" ~ "Deceased"),
         BPD.Grade=case_when(jensen=="grado 1" ~ "Grade 1",
                             jensen=="grado 2" ~ "Grade 2",
                             jensen=="grado 3" ~ "Grade 3",
                             jensen=="control" ~ "Control",
                             jensen=="fallecido" ~ "Deceased"),
         Days.hospitalization=difftime(`Fecha de alta`,`Fecha de Nacimiento`, 
                                       units = "days"),
         New.Sex=case_when(Sexo=="Mujer" ~ "Yes",
                           Sexo=="Varón" ~ "No"))  %>%
  mutate_all(function(x) gsub("Varón","Male",x)) %>%
  mutate_all(function(x) gsub("Mujer","Female",x))

variables <- variables %>%
  mutate_at(c("EG al nacimiento (semanas)", "Peso al nacimiento", 
              "Days.hospitalization", "Días 1 sem de VM"), as.numeric)
variables$Condition <- factor(variables$Condition, levels=c("Control", 
                                                            "BPD",
                                                            "Deceased"))
variables$BPD.Grade <- factor(variables$BPD.Grade, levels=c("Control", "Grade 1",
                                                            "Grade 2", "Grade 3",
                                                            "Deceased"))
variables$New.Sex <- factor(variables$New.Sex, levels = c("Yes", "No"))
wide.modality <- read.xlsx("wide_060324_mod.xlsx")
wide.modality <- wide.modality %>%
  mutate(Modality=case_when(Modalidad=="AC/VG" | Modalidad=="PC" | 
                              Modalidad=="PSV" | Modalidad=="PSV+VG" |
                              Modalidad=="SIMV+VG" | Modalidad=="VAFO" |
                              Modalidad=="VCRP" | Modalidad=="VM A/C" |
                              Modalidad=="VM A/C CON VG" | 
                              Modalidad=="VMC" | Modalidad=="VMC A/C" ~ "IMV",
                            Modalidad=="VNI" | Modalidad=="DUOPAP" |
                              Modalidad=="IPPVn" ~ "NIV",
                            Modalidad=="CPAP" | Modalidad=="CPAPn" ~ "nCPAP",
                            Modalidad=="Alto flujo" | Modalidad=="GNAF" |
                              Modalidad=="OAF" | Modalidad=="HFNC" ~ "HFNC",
                            Modalidad=="GNBF" ~ "LFNC",
                            Modalidad=="No O2" | 
                              Modalidad=="Sin soporte" ~ "No respiratory support"
                            ))
wide.modality$Modality <- factor(wide.modality$Modality, levels = c("IMV", "NIV",
                                                                    "nCPAP",
                                                                    "HFNC","LFNC",
                                                                    "No respiratory support"))
new.variables <- merge(variables, wide.modality, by=c("patient","cod_centro"))
new.variables <- new.variables[,c(16,19,12:14,21,17:18)]

new.variables %>%
  tbl_summary(
    by= Condition,
    statistic = list(all_continuous() ~ "{mean} ({sd})",        
                     BPD.Grade ~ "{n} ({p}%)",
                     New.Sex ~ "{n} ({p}%)"),   
    digits = all_continuous() ~ 1,                            
    type   = list(all_categorical() ~ "categorical",
                  `EG al nacimiento (semanas)` ~ "continuous",
                  `Días 1 sem de VM` ~ "continuous",
                  Modality ~ "categorical"),
    label = list(New.Sex ~ "Sex, Female",
                 `EG al nacimiento (semanas)` ~ "GA (weeks)",
                 `Peso al nacimiento` ~ "Weight",
                 `Días 1 sem de VM` ~ "Days of VM at 1 week",
                 BPD.Grade ~ "BPD grade",
                 Days.hospitalization ~ "Days of hospitalization",
                 Modality ~ "Respiratory modality al 1 week")) %>%
  modify_header(label ~ "**Condition**") %>%
  modify_caption("**Table 1. Patient Clinical Characteristics**") %>%
  add_p() %>%
  bold_p() %>%
  add_overall() %>%
  italicize_levels() %>%
  as_gt() %>%
  gt::tab_source_note(gt::md("*IVM: Invasive Mechanical Ventilation; NIV: Non-Invasive Ventilation; nCPAP: nasal CPAP; HFNC: High Flow Nasal Cannula ; LFNG: Low Flow Nasal Cannula*"))

new.variables %>%
  tbl_summary(
    by= Condition,
    statistic = list(all_continuous() ~ "{mean} ({sd})",        
                     BPD.Grade ~ "{n} ({p}%)",
                     New.Sex ~ "{n} ({p}%)"),   
    digits = all_continuous() ~ 1,                            
    type   = list(all_categorical() ~ "categorical",
                  `EG al nacimiento (semanas)` ~ "continuous",
                  `Días 1 sem de VM` ~ "continuous",
                  Modality ~ "categorical"),
    label = list(New.Sex ~ "Sex, Female",
                 `EG al nacimiento (semanas)` ~ "GA (weeks)",
                 `Peso al nacimiento` ~ "Weight",
                 `Días 1 sem de VM` ~ "Days of VM at 1 week",
                 BPD.Grade ~ "BPD grade",
                 Days.hospitalization ~ "Days of hospitalization",
                 Modality ~ "Respiratory modality al 1 week")) %>%
  modify_header(label ~ "**Condition**") %>%
  modify_caption("**Table 1. Patient Clinical Characteristics**") %>%
  add_p() %>%
  bold_p() %>%
  add_overall() %>%
  italicize_levels() %>%
  as_gt() %>%
  gt::tab_source_note(gt::md("*IVM: Invasive Mechanical Ventilation; NIV: Non-Invasive Ventilation; nCPAP: nasal CPAP; HFNC: High Flow Nasal Cannula ; LFNG: Low Flow Nasal Cannula*")) %>%
  gt::gtsave(filename = "../Table_1_with_DEC.docx")

```

